Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-01 demystify.tex
updates

Additionally, \textbf{work experiences at renowned research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicit say something about them in your statement.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than 
Masters.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have higher "activity" or some other hidden fee (e.g., you may need to pay some fees for each credit hour).

Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In most cases, it is also enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  At GMU CS, students admitted with TA have  4 years of GTA guaranteed, and in some cases also receive  stipend for the first summer.
\end{commentbox}

@@

\textbf{How to get RA?} When a professor recruits you, they will likely give you RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an adviser to support you with RA. See how to contact a prof. in \autoref{sec:contact} for research opportunities.

\begin{commentbox}[Vu:]
  If you got recruited by a prof., who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD adcom, wants to work with and funds you, the PhD adcom will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

@@

In general, fellowship, especially competitive ones from the government or company, is prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only superstars have NSF grad or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even ``honorable mentioning'', school will still mention you on their website and you still should put it on your CV.

@@

\section{Why did I get rejected?}
Many possible reasons!  Other than the obvious ones (e.g., weak research pontentials, low GPA, weak LoRs, unclear SOP), here are some that you might not expect:
\begin{itemize}
  \item  \emph{Your English scores (e.g., IELTS or TOEFL) are not high enough}.  While professors do not care much about these, there is a certain minimum set by the college or the university that you need to pass to get admitted, especially if you want to get funding (e.g., you need to be able to communicate well with students as a TA).  Note this is mostly for international students, as domestic one do not have to take these tests.

@@

\paragraph{Should I ask for feedback?}
No, don't bother.  You will likely not get any useful feedback.  We cannot reveal your evaluation results (e.g., if we say because you have weak LORs, then that would reveal something about your writers) or give you advice on how to improve. So it is best to just move on.  If you really want feedback, ask your professors, collaborators, ref writers, or those who have previously applied.

\section{What can you do to increase your admission chance?}\label{sec:improve-your-chance}

@@

Show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for "URM minority in CS in the US" on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

>>>>

Additionally, \textbf{work experiences at well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicit say something about them in your statement.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhDs are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than 
Masters.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have extra "activity" fee or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

\paragraph{Low Stipend?} Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In most cases, it is also enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  At GMU CS, PhD students admitted TA has a 4 years of GTA guaranteed, and in also receive  stipend for the first summer.
\end{commentbox}

@@

\textbf{How to get RA?} When a professor recruits you, they might offer you an RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an adviser to support you with RA. See how to contact a prof. in \autoref{sec:contact} for research opportunities.

\begin{commentbox}[Vu:]
  If you got recruited an offered an RA by a prof., it's very likely you will get admitted.  For example, if a prof., even if not in PhD adcom, wants to fund you, the PhD adcom will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

@@

In general, fellowship, especially competitive ones from the government or company, is prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only few  would have NSF Graduate or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even ``honorable mentioning'', school will still mention you on their website and you still should put it on your CV.

@@

\section{Why did I get rejected?}
Many possible reasons!  Other than having weak profile (e.g., research potentials, GPA, LoRs, SOP), here are some that you might not expect:
\begin{itemize}
  \item  \emph{Your English scores (e.g., IELTS or TOEFL) are not high enough}.  While professors might not care much about these, there is a certain minimum set by the college or the university that you need to pass to get admitted, especially for TA funding (e.g., you need to be able to communicate well with students as a TA).  Moreover, while professors are willing to fight to get you admitted, they might be reluctant to go against the standard requirement of the university or college on English scores.  Note this is mostly for international students, as domestic one do not have to take these tests.

@@

\paragraph{Should I ask for feedback?}
No, don't bother.  You will likely not get any useful feedback.  We are not willing to reveal your evaluation results or give you advice on how to improve. So it is best to just move on.  If you really want feedback, ask your professors, collaborators, ref writers, or those who have previously applied.

\section{Increasing your admission chance}\label{sec:improve-your-chance}

@@

Given the high number of quality applicants and limited number of spots, other than having a good application profile, you really want to show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for "URM minority in CS in the US" on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.
